---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# kmetadata

<!-- badges: start -->
<!-- badges: end -->

The **kmetadata** package provides tools for managing and manipulating
metadata in R. It simplifies the process of creating, editing, and
loading metadata for datasets, making it easier to document variable
names, labels, and value definitions.

## Installation

You can install the development version of kmetadata from GitLab:

``` r
# install.packages("remotes")
remotes::install_gitlab("kbosirany/kmetadata")
```

## Features

- **Create metadata templates** for variable names and unique values
- **Support for multiple data formats**: data.frames, lists, and files
- **Excel integration**: Export metadata to Excel for manual editing
- **Automatic caching**: Load existing metadata or create new templates
- **Flexible workflow**: Force recreation or use cached versions

## Basic Usage

### Creating Variable Metadata

Create metadata templates for documenting variable names:

```{r example-vars}
library(kmetadata)

# Sample dataset
survey_data <- data.frame(
  participant_id = 1:5,
  age_years = c(25, 30, 35, 28, 42),
  gender = c("M", "F", "M", "F", "M"),
  satisfaction = c("High", "Medium", "High", "Low", "Medium")
)

# Create variable metadata
var_metadata <- create_metadata(
  survey_data,
  path = "survey.csv",
  which = "vars"
)

# View the structure
str(var_metadata)
```

The metadata includes columns for:
- `label`: Original variable name
- `new_label`: Space for descriptive labels
- `description`: Space for detailed descriptions

### Creating Value Metadata

Document unique values within categorical variables:

```{r example-values}
# Create value metadata for specific variables
value_metadata <- create_metadata(
  survey_data,
  path = "survey.csv",
  which = "values",
  vars = c("gender", "satisfaction")
)

# View gender metadata
value_metadata$gender
```

### Writing Metadata to Excel

Export metadata templates to Excel for manual editing:

```{r example-write, eval=FALSE}
# Create and write metadata to Excel file
create_metadata(
  survey_data,
  path = "data/survey.csv",
  which = "vars",
  write_results = TRUE
)

# This creates: data/survey - vars - metadata.xlsx
```

### Loading or Creating Metadata

Use `load_metadata()` to automatically load existing metadata or
create new templates:

```{r example-load, eval=FALSE}
# First time: creates metadata
metadata <- load_metadata(
  path = "data/survey.csv",
  which = "vars",
  vars = c("participant_id", "age_years", "gender", "satisfaction")
)

# After manual editing in Excel, this loads the edited version
metadata <- load_metadata(
  path = "data/survey.csv",
  which = "vars",
  vars = c("participant_id", "age_years", "gender", "satisfaction")
)

# Force recreation (ignores cached version)
metadata <- load_metadata(
  path = "data/survey.csv",
  which = "vars",
  vars = c("participant_id", "age_years", "gender", "satisfaction"),
  force = TRUE
)
```

### Working with Multiple Sheets

Create metadata for multi-sheet Excel files:

```{r example-multisheet, eval=FALSE}
# Data with multiple sheets
data_list <- list(
  demographics = data.frame(
    id = 1:3,
    age = c(25, 30, 35),
    gender = c("M", "F", "M")
  ),
  responses = data.frame(
    id = 1:3,
    q1 = c("Yes", "No", "Yes"),
    q2 = c("A", "B", "A")
  )
)

# Create variable metadata for all sheets
var_meta <- create_metadata(
  data_list,
  path = "survey_data.xlsx",
  which = "vars",
  write_results = TRUE
)

# Create value metadata for specific variables in each sheet
val_meta <- create_metadata(
  data_list,
  path = "survey_data.xlsx",
  which = "values",
  vars = list(
    demographics = c("gender"),
    responses = c("q1", "q2")
  ),
  write_results = TRUE
)
```

## Workflow Example

1. **Create initial metadata template**:
```{r workflow-1, eval=FALSE}
create_metadata(
  "data/survey.csv",
  which = "vars",
  write_results = TRUE
)
```

2. **Edit the Excel file manually** to add labels and descriptions

3. **Load the edited metadata**:
```{r workflow-2, eval=FALSE}
metadata <- load_metadata(
  path = "data/survey.csv",
  which = "vars",
  vars = colnames(survey_data)
)
```

4. **Use metadata in your analysis**:
```{r workflow-3, eval=FALSE}
# Access descriptive labels
metadata$vars$new_label

# Access descriptions
metadata$vars$description
```

## File Naming Convention

Metadata files are automatically named following this pattern:

```
[original_filename] - [type] - metadata.xlsx
```

Examples:
- `survey.csv` → `survey - vars - metadata.xlsx`
- `survey.csv` → `survey - values - metadata.xlsx`
- `data.xlsx` → `data - vars - metadata.xlsx`

## Getting Help

For more information on specific functions:

```{r help, eval=FALSE}
?create_metadata
?load_metadata
```

## License

AGPL (>= 3)

## Author

Kevin Bosirany Orlando (kevinbosirany@gmail.com)
ORCID: 0009-0009-2784-3108
