---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# kmetadata

<!-- badges: start -->
<!-- badges: end -->

The **kmetadata** package provides tools for creating and managing
metadata for R datasets. It simplifies documenting variable names,
labels, and value definitions through an Excel-based workflow.

## Installation

Install the development version from GitHub:

``` r
# install.packages("remotes")
remotes::install_github("kbosirany/kmetadata")
```

## Quick Start

```{r example}
library(kmetadata)

# Sample data
data <- data.frame(
  id = 1:3,
  gender = c("M", "F", "M"),
  status = c("Active", "Inactive", "Active")
)

# Create variable metadata
var_meta <- create_metadata(data, path = "data.csv", which = "vars")
var_meta

# Create value metadata for categorical variables
val_meta <- create_metadata(data, path = "data.csv", which = "values",
                            vars = c("gender", "status"))
val_meta$gender
```

## Key Features

- **Create metadata templates** for variables and values
- **Export to Excel** for manual editing and documentation
- **Smart caching** with `load_metadata()` - loads existing or creates
  new
- **Multi-sheet support** for Excel workbooks
- **Simple workflow** - create, edit, load

## Typical Workflow

1. **Create metadata template**:
```{r workflow-1, eval=FALSE}
create_metadata("data/survey.csv", which = "vars",
                write_results = TRUE)
```

2. **Edit in Excel** - add labels and descriptions

3. **Load edited metadata**:
```{r workflow-2, eval=FALSE}
metadata <- load_metadata(path = "data/survey.csv",
                          which = "vars",
                          vars = colnames(data))
```

## Documentation

- **Vignette**: `vignette("kmetadata-intro")` - Comprehensive guide
  with examples
- **Functions**: `?create_metadata`, `?load_metadata`
- **File naming**: Outputs follow pattern
  `[filename] - [type] - metadata.xlsx`

## Learn More

See the package vignette for detailed examples including:

- Variable and value metadata creation
- Working with multi-sheet Excel files
- Custom file paths and options
- Complete workflow examples
- Tips and best practices

## License

AGPL (>= 3)

## Author

Kevin Bosirany Orlando  
Email: kevinbosirany@gmail.com  
ORCID: [0009-0009-2784-3108](https://orcid.org/0009-0009-2784-3108)
