---
title: "Introduction to kmetadata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to kmetadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(kmetadata)
```

## Overview

The **kmetadata** package provides tools for managing and manipulating
metadata in R. It simplifies the process of creating, editing, and
loading metadata for datasets, making it easier to document variable
names, labels, and value definitions.

## Key Features

- **Create metadata templates** for variable names and unique values
- **Support for multiple data formats**: data.frames, lists, and files
- **Excel integration**: Export metadata to Excel for manual editing
- **Automatic caching**: Load existing metadata or create new templates
- **Flexible workflow**: Force recreation or use cached versions

## Basic Concepts

### Metadata Types

The package supports two types of metadata:

1. **Variable metadata (`which = "vars"`)**: Documents variable names
   with labels and descriptions
2. **Value metadata (`which = "values"`)**: Documents unique values
   within categorical variables

### File Naming Convention

Metadata files are automatically named following this pattern:

```
[original_filename] - [type] - metadata.xlsx
```

Examples:
- `survey.csv` → `survey - vars - metadata.xlsx`
- `survey.csv` → `survey - values - metadata.xlsx`
- `data.xlsx` → `data - vars - metadata.xlsx`

## Creating Variable Metadata

Variable metadata helps document the meaning and purpose of each
variable in your dataset.

```{r example-vars}
# Sample dataset
survey_data <- data.frame(
  participant_id = 1:5,
  age_years = c(25, 30, 35, 28, 42),
  gender = c("M", "F", "M", "F", "M"),
  satisfaction = c("High", "Medium", "High", "Low", "Medium")
)

# Create variable metadata
var_metadata <- create_metadata(
  survey_data,
  path = "survey.csv",
  which = "vars"
)

# View the structure
str(var_metadata)
```

The metadata template includes three columns:

- `label`: Original variable name
- `new_label`: Space for descriptive labels (initially NA)
- `description`: Space for detailed descriptions (initially NA)

## Creating Value Metadata

Value metadata documents the unique values within categorical variables,
useful for creating codebooks and data dictionaries.

```{r example-values}
# Create value metadata for specific variables
value_metadata <- create_metadata(
  survey_data,
  path = "survey.csv",
  which = "values",
  vars = c("gender", "satisfaction")
)

# View gender metadata
value_metadata$gender

# View satisfaction metadata
value_metadata$satisfaction
```

## Writing Metadata to Excel

Export metadata templates to Excel files for manual editing by
colleagues or collaborators.

```{r example-write, eval=FALSE}
# Create and write metadata to Excel file
create_metadata(
  survey_data,
  path = "data/survey.csv",
  which = "vars",
  write_results = TRUE
)

# This creates: data/survey - vars - metadata.xlsx
```

You can then open this file in Excel and fill in:
- **new_label**: Descriptive, human-readable labels
- **description**: Detailed information about each variable

### Overwriting Existing Files

By default, the function won't overwrite existing metadata files.
Use `overwrite = TRUE` to replace them:

```{r overwrite, eval=FALSE}
create_metadata(
  survey_data,
  path = "data/survey.csv",
  which = "vars",
  write_results = TRUE,
  overwrite = TRUE  # Force overwrite
)
```

## Loading Metadata

The `load_metadata()` function provides an intelligent workflow:

1. **First time**: Creates new metadata if none exists
2. **Subsequent times**: Loads existing metadata from Excel
3. **With `force = TRUE`**: Recreates metadata, ignoring cached version

```{r example-load, eval=FALSE}
# First time: creates metadata
metadata <- load_metadata(
  path = "data/survey.csv",
  which = "vars",
  vars = c("participant_id", "age_years", "gender", "satisfaction")
)

# After manual editing in Excel, this loads the edited version
metadata <- load_metadata(
  path = "data/survey.csv",
  which = "vars",
  vars = c("participant_id", "age_years", "gender", "satisfaction")
)

# Access the metadata
metadata$vars$new_label
metadata$vars$description
```

### Forcing Metadata Recreation

Use `force = TRUE` to recreate metadata even if a file exists:

```{r force, eval=FALSE}
# Force recreation (ignores cached version)
metadata <- load_metadata(
  path = "data/survey.csv",
  which = "vars",
  vars = c("participant_id", "age_years", "gender", "satisfaction"),
  force = TRUE
)
```

## Working with Files

You can create metadata directly from file paths:

```{r file-example, eval=FALSE}
# Create metadata from a CSV file
metadata <- create_metadata(
  "data/survey.csv",
  which = "vars",
  write_results = TRUE
)

# Create value metadata from a file
value_meta <- create_metadata(
  "data/survey.csv",
  which = "values",
  vars = c("gender", "satisfaction"),
  write_results = TRUE
)
```

## Working with Multiple Sheets

For Excel files with multiple sheets, the package handles each sheet
separately.

```{r example-multisheet, eval=FALSE}
# Data with multiple sheets
data_list <- list(
  demographics = data.frame(
    id = 1:3,
    age = c(25, 30, 35),
    gender = c("M", "F", "M")
  ),
  responses = data.frame(
    id = 1:3,
    q1 = c("Yes", "No", "Yes"),
    q2 = c("A", "B", "A")
  )
)

# Create variable metadata for all sheets
var_meta <- create_metadata(
  data_list,
  path = "survey_data.xlsx",
  which = "vars",
  write_results = TRUE
)

# Examine the structure
names(var_meta)  # "demographics" "responses"
var_meta$demographics
var_meta$responses
```

### Value Metadata for Multiple Sheets

You can specify which variables to document in each sheet:

```{r multisheet-values, eval=FALSE}
# Create value metadata for specific variables in each sheet
val_meta <- create_metadata(
  data_list,
  path = "survey_data.xlsx",
  which = "values",
  vars = list(
    demographics = c("gender"),
    responses = c("q1", "q2")
  ),
  write_results = TRUE
)

# Names include both sheet and variable
names(val_meta)  # "demographics.gender" "responses.q1" "responses.q2"

# View specific metadata
val_meta$demographics.gender
val_meta$responses.q1
```

## Complete Workflow Example

Here's a typical workflow for documenting a dataset:

### Step 1: Create Initial Metadata Template

```{r workflow-1, eval=FALSE}
# Create metadata template and export to Excel
create_metadata(
  "data/survey.csv",
  which = "vars",
  write_results = TRUE
)

# Also create value metadata for categorical variables
create_metadata(
  "data/survey.csv",
  which = "values",
  vars = c("gender", "satisfaction", "education"),
  write_results = TRUE
)
```

### Step 2: Edit Metadata in Excel

Open the generated Excel files and fill in:
- Descriptive labels in the `new_label` column
- Detailed descriptions in the `description` column

Example edits:
- `participant_id` → "Participant ID" / "Unique identifier for each
  survey respondent"
- `age_years` → "Age (years)" / "Respondent's age at time of survey"
- `gender` value "M" → "Male" / "Male respondent"

### Step 3: Load the Edited Metadata

```{r workflow-3, eval=FALSE}
# Load the edited metadata
var_metadata <- load_metadata(
  path = "data/survey.csv",
  which = "vars",
  vars = colnames(survey_data)
)

value_metadata <- load_metadata(
  path = "data/survey.csv",
  which = "values",
  vars = c("gender", "satisfaction", "education")
)
```

### Step 4: Use Metadata in Analysis

```{r workflow-4, eval=FALSE}
# Access descriptive labels
var_metadata$vars$new_label

# Access descriptions for documentation
var_metadata$vars$description

# Create a data dictionary
data_dict <- var_metadata$vars
print(data_dict)

# Use value metadata for recoding
gender_labels <- setNames(
  value_metadata$gender$new_label,
  value_metadata$gender$label
)
```

## Selecting Specific Variables

You can limit metadata to specific variables using the `vars` parameter:

```{r select-vars}
# Create metadata for only specific variables
selected_meta <- create_metadata(
  survey_data,
  path = "survey.csv",
  which = "vars",
  vars = c("participant_id", "gender")
)

# Note: the metadata still includes all variables from the data frame
# Use vars to control which value metadata is created
value_meta <- create_metadata(
  survey_data,
  path = "survey.csv",
  which = "values",
  vars = "gender"  # Only gender values
)

names(value_meta)  # Only "gender"
```

## Custom Metadata File Paths

By default, metadata files are created in the same directory as the
data file. You can specify a custom path:

```{r custom-path, eval=FALSE}
# Custom metadata file path
create_metadata(
  survey_data,
  path = "data/survey.csv",
  which = "vars",
  path_metadata = "metadata/custom_name.xlsx",
  write_results = TRUE
)
```

## Tips and Best Practices

1. **Start early**: Create metadata templates at the beginning of your
   project
2. **Version control**: Track metadata files alongside your data
3. **Consistent naming**: Use the default naming convention for clarity
4. **Document thoroughly**: Take time to write clear descriptions
5. **Update regularly**: Refresh metadata when variables change
6. **Use value metadata**: Essential for categorical variables and
   codebooks
7. **Backup**: Keep copies of manually edited metadata files

## Common Use Cases

### Creating a Data Dictionary

```{r data-dict, eval=FALSE}
# Create complete documentation
var_meta <- create_metadata("data.csv", which = "vars",
                             write_results = TRUE)
val_meta <- create_metadata("data.csv", which = "values",
                             vars = c("status", "category", "type"),
                             write_results = TRUE)

# After editing in Excel, load and combine
variables <- load_metadata("data.csv", which = "vars", 
                            vars = colnames(data))
values <- load_metadata("data.csv", which = "values",
                         vars = c("status", "category", "type"))

# Create formatted data dictionary
dictionary <- list(
  variables = variables$vars,
  values = values
)
```

### Sharing Data with Collaborators

```{r sharing, eval=FALSE}
# Create comprehensive metadata
create_metadata("shared_data.csv", which = "vars", 
                write_results = TRUE)
create_metadata("shared_data.csv", which = "values",
                vars = c("treatment", "outcome", "site"),
                write_results = TRUE)

# Share both the data file and metadata Excel files
# Collaborators can understand the data structure immediately
```

### Documenting Survey Data

```{r survey, eval=FALSE}
# Document all survey questions
create_metadata("survey_responses.csv", which = "vars",
                write_results = TRUE)

# Document response options
likert_vars <- c("q1_satisfaction", "q2_agreement", "q3_importance")
create_metadata("survey_responses.csv", which = "values",
                vars = likert_vars, write_results = TRUE)
```

## Further Reading

- `?create_metadata` - Detailed function documentation
- `?load_metadata` - Loading and caching metadata
- Package README - Quick start guide
